---
title: "Gapminder Exploration"
author: "Haohui Zhong"
date: "2019/9/17"
output: ioslides_presentation
---
```{r load,warning=FALSE,echo=FALSE}
library(gapminder)
library(tibble)
library(DT)
```

## Chapters

<br>

1. Outline of dataset `gapminder`

2. Data analysis

+ Maximum, minimun, mean, range

+ Filtering data

+ Visualizing data


## An Outline of dataset `gapminder`

- **a general description of gapminder**
```{r,echo=FALSE,warning=FALSE}
summary(gapminder)

```

- **Some basic information{.smaller}**

How many **rows** does gapminder have?

```{r,echo=FALSE,warning=FALSE}
nrow(gapminder)

```

How many **columns** does gapminder have?

```{r,echo=FALSE,warning=FALSE}
ncol(gapminder)

```

What variables are in this dataset?

```{r, comment = NA}
colnames(gapminder)
```

## Data analysis

> Maximum, minimun, mean, range

Take `GDP per capita` as an example:

```{r,echo=FALSE,warning=FALSE}
summary(gapminder$gdpPercap)
```

## Data analysis {.smaller}
> Filtering data

Suppose that we are carrying out **country-specific** studies and interested in **Australia**. We need to break down the dataset and collect observations of interest.
```{r,echo=FALSE,warning=FALSE}
gapminder[gapminder$country == "Australia", ]
```

## Data analysis {.smaller}

> Visualizing the dataset

The graphic below shows how GDP per capita of all countries changes over time.

```{r, echo=FALSE,warning=FALSE}
plot(gdpPercap ~ year, gapminder)
```

## 
<div align="center">

Thanks for your attention



