---
title: "hw03-dplyr-ggplot2"
author: "Haohui Zhong"
date: "2019/9/30"
output: html_document
---

# Outline

This assignment is an extension version which takes us to a higher level exploration of datasets such as `gapminder`. Utilizing `dplyr` and `ggplot2`, we are about to accomplish the goal of analyzing and visualizing the dataset. As usual, we need to load packages in the very beginning.

```{r load, message = FALSE}
library(gapminder)
library(tidyverse)
library(scales)
```

Now equipped with proper tools, we are ready to get familiar with them by finishing the following tasks.

# Task 1

_Requirements:_
Report the absolute or relative abundance of countries with low life expectancy over time by continent. Then determine how many countries on each continent have a life expectancy less than the chosen benchmark, for each year.

Here we define life expectancy less than the mean value of life expectancy across the world as `low` life expectancy.


```{r}
gapminder %>% 
  group_by(year) %>% 
  filter(lifeExp < mean(lifeExp)) %>% 
  group_by(year, continent) %>% 
  summarise(lowlifeExpCount = length(continent))

```

The above chunk groups data by `year` and `continent`. Then compares life expectancy of each country in each year with the chosen benchmark, that is the `mean value` for the corresponding year and counts the number of countries with low life expectancy within continents for each year.

By visualizing the above results, we can present them in a better way.

```{r}
# plotting
gapminder %>% 
  group_by(year) %>% 
  mutate(mean_lifeExp = mean(lifeExp)) %>% 
  group_by(continent) %>%
  filter(lifeExp < mean_lifeExp) %>% 
  group_by(year) %>% 
  count(continent) %>% 
  ggplot(aes(year,n, color = continent))+
  geom_line(size = 1)+
  xlab("Year")+
  ylab("Number of lowlifeExp Countries")+
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 14))

```

From the graph, we can clearly see that Africa accounts for the largest proportion of low life expectancy countries. And neither countries in Oceania have had low life expectancy ever since the year of observation, nor countries in Europe have had since the late 1980s. There is a common decreasing trend in numbers of low life expectancy countries among all continents.

The above chunks show us the absolute abundance of low life expectancy countries and now we are going to study the relative one, which focuses on the relationship between numbers of low life expectancy countries within a continent and the total amount of them.

```{r}
# plotting
gapminder %>% 
  group_by(year) %>% 
  mutate(mean_lifeExp = mean(lifeExp)) %>% 
  group_by(year, continent) %>% 
  mutate(total = n()) %>% 
  filter(lifeExp < mean_lifeExp) %>%
  group_by(year, continent) %>% 
  mutate(cc = n()) %>%
  select(continent, year, cc, total) %>% 
  mutate(ra = (cc / total)*100) %>% 
  distinct() %>%
  ggplot(aes(year, ra, fill = continent))+
  geom_bar(stat = "identity")+
  xlab("Year")+
  ylab("Relative abundance(%)")

```

By studying absolute abundance, we learn exact values of each continent for each year, while working on relative abundance shows variations among continents within the same year.

# Task 2

_Requirements:_
Obtain the maximum and minimum of GDP per capita for all continents.

This task seems to be quite simple in the very beginning. All we have to do is to group the data and compute minimum and maximum values of GDP per capita within continents.

```{r}
gapminder %>%
  group_by(continent) %>%
  summarise(min = min(gdpPercap), max = max(gdpPercap))
t3 = gapminder
```

The above table shows minimum and maximum values of GDP per capita within each continent over the years. We can further explore the GDP data by studying ranges of GDP per capita of each continent and compare these ranges with one another, etc. 

```{r}
# plotting
t3 %>%
  ggplot(aes(continent, gdpPercap, color = continent))+
  geom_boxplot(size = 1.5)+
  scale_y_log10(labels = scales::dollar_format()) +
  stat_summary(aes(label = round(10^..y..)), fun.y = min, geom = "text", size = 3, vjust = 1.5) +
  stat_summary(aes(label = round(10^..y..)), fun.y = max, geom = "text", size = 3, vjust = -1) +
  xlab("Continent") +
  ylab("GDP per capita")

```

We can learn that both maximum and minimum values of Africa rank last among continents, and that Asia has an extremely unequal distribution of GDP per capita while distribution in Oceania is very intensive since there are only two countries.

# Task 5

_Requirements:_
Find out how life expectancy changes over time on different continents.

Actually we've already gone through this part in `assignment 2`. This is how we did it in `assignment 2`.

```{r}
gapminder %>% 
  group_by(continent) %>% 
  ggplot(aes(color = continent,x = year, y = lifeExp))+
  geom_point(alpha = 0.4) +
  xlab("year")+
  ylab("life expectancy")
```

We can barely see some trends, but this is not very efficient when presenting data. So we can make some improvements by choosing mean values of life expectancy of each continent as representatives of the corresponding continent, instead of looking at all countries within the continent.

```{r}
# plotting
gapminder %>% 
  group_by(continent,year) %>% 
  mutate(meanlifeExp = mean(lifeExp)) %>% 
  ggplot(aes(color = continent,x = year, y = meanlifeExp))+
  geom_line(size = 1) +
  xlab("year")+
  ylab("mean of life expectancy")

```

If we look at the point plot, we might draw the conclusion that Europe enjoys an extremely long life expectancy, followed by Americas. But if we use the line graph, we will see that Oceania actually has the longest life expectancy. We would probably neglect data of Oceania in a point plot due to the extremely limited number of it.

What if we also take population into consideration?

Things would be a little bit different.

```{r}
# plotting
gapminder %>% 
  group_by(continent,year) %>% 
  mutate(meanlifeExp = weighted.mean(lifeExp,pop)) %>% 
  ggplot(aes(color = continent,x = year, y = meanlifeExp))+
  geom_line(size = 1) +
  xlab("year")+
  ylab("weighted mean of life expectancy")


```

Such results remind us of the importance of choosing a proper measurement when conducting researches.








